Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2022-01-25 document/source/body.tex
Shortened Section 4F
Shortened Section 4F on Actor Navigation by 50%.

In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.

To briefly describe the A* process, at each stage of the search, the algorithm selects one node for expansion from a set of nodes known as the \textit{frontier} (also fringe or open set) using a heuristic to select this node. For reasons of efficiency, A* represents the frontier as a priority queue, with nodes prioritised by the heuristic. Selection then reduces to a dequeue operation on the frontier. In general, the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered. The question of how these ties are broken in a priority queue is related to the broader notion of sorting stability~\cite{Sedgewick2011}. An implementation of a priority queue is said to be stable if it preserves insertion order, otherwise it is said to be unstable. If an implementation of A* uses a stable priority queue, then it will behave as a deterministic algorithm which always returns the optimal node that was first added to the frontier. Stable priority queues are typically less computationally efficient than unstable priority queues, so the requirement on stability is often dropped in practice.

For example, one of the most common implementations of a priority queue relies on a binary heap and is unstable; it will always return the first optimal node according to its internal heap order rather than insertion order. The default priority queue implementations in the standard Java and Boost libraries are based on binary heaps~\cite{boost2021,priorityQueue}, while the Unreal documentation and source code suggests that this is also true of CARLA~\cite{FGraphAStar,TArray,GraphAStar}. Thus, an unstable but deterministic priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order. If an implementation of A* uses an unstable but deterministic priority queue, then it will behave as a deterministic algorithm which always returns the optimal node according the ordering of its priority queue.

Finally, an unstable but deterministic priority queue suggests a third kind of priority queue that we can refer to as an unstable and non-deterministic priority queue; that is, a priority queue that always returns an optimal node selected at random (i.e. breaking ties arbitrarily). If an implementation of A* uses an unstable and non-deterministic priority queue, then it will behave as a non-deterministic algorithm. Overall, an implementation of A* is much more likely to use a priority queue that is unstable but deterministic rather than one that is stable, and is very unlikely to use one that is unstable and non-deterministic.

An implementation of A* will behave as a deterministic algorithm if it uses a stable priority queue or an unstable but deterministic priority queue. However, apparent determinism of A* will be sensitive to the ordering preserved by the priority queue (i.e. insertion order or e.g. heap order). If we are not aware of changes to this ordering, then a deterministic implementation of A* may appear to behave as a non-deterministic algorithm, since we will not have accounted for changes to the ordering. For example, suppose actions are iterated over in order of hash value (where order of iteration over actions often determines insertion order), but between runs there is some untracked change to the software/hardware stack that changes the calculation of the hash value, then this may change insertion order of nodes to the frontier, which could eventually materialise as a different optimal solution found by A*. This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator. The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.

To summarise, documentation suggests that CARLA uses a binary heap for the priority queue~\cite{FGraphAStar,TArray,GraphAStar} which preserves some order of the frontier nodes, but not necessarily the insertion order, and hence is unstable. Therefore, determinism of the A* algorithm for actor navigation will be sensitive to the ordering preserved by the priority queue, in this case the heap order. This also assumes the simulation environment itself is deterministic as sources of non-determinism that affect the simulator or the system more generally may cause changes to the operation of an A* implementation.

>>>>

The A* algorithm is commonly used for actor navigation in game engines.
A* explores a search graph through a series of node-selection-and-expansion steps that continue until a goal node is found.
A heuristic is used at each step to select the most promising node for expansion from a set of candidate nodes called the \textit{frontier}.
The heuristic is not required to guarantee a uniquely preferred node in the frontier, so ties may be found during selection and then broken arbitrarily, meaning that an implementation of A* can be non-deterministic.

For reasons of efficiency, A* represents the frontier as a priority queue with nodes prioritised by the heuristic.
Selection then reduces to a dequeue operation on the frontier.
How to break ties in priority queues is related to a broader question of sorting stability~\cite{Sedgewick2011} where a priority queue is stable if it breaks ties based on insertion order, and is unstable otherwise.
Typically unstable priority queues are more computationally efficient than stable priority queues and the most common method (using a binary heap)~\cite{boost2021,priorityQueue} is unstable, breaking ties based on internal heap order rather than insertion order.
While this priority queue is unstable, it is still deterministic, since it always breaks ties in the same way (based on heap order).
An example of an unstable priority queue that is non-deterministic is one that breaks ties at random.
An implementation of A* is deterministic if it uses either a stable priority queue or an unstable but deterministic priority queue.

The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap~\cite{FGraphAStar,TArray,GraphAStar}.
Theoretically this means that use of A* in CARLA will be determinstic as long as the heap is always constructed in the same way.
However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may then appear non-deterministic.
This can be a complex problem in practice.
For example, suppose A* iterates over actions in order of hash value, but between runs there is some untracked change to the hardware/software stack that alters the hash function, then this may change insertion order of nodes to the frontier, which could eventually lead to a different binary heap and thus a different optimal solution found by A*.
This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator.
The difference between use of a stable priority queue and an unstable but deterministic priority is simply that it may be easier to detect such changes under the former because insertion order is typically more meaningful with respect to the actual implementation of A*.
